An agent-based model of military mechanization

### 2018-05-13 paper2/paper2.tex
foo

\begin{centering}
	Word count: XXXX
\end{centering}
\section{Introduction}

@@

Agents exist in a spatial environment. Spatiality is represented by neighbor
relationships. Two states are defined as
being neighbors if they share a contiguous land border or a stretch of water
less than 400 miles. Data on neighbor relationships is taken from the Correlates
of War (COW) Direct Continuity dataset (\cite{cow_contiguity}).
Modeling spatiality as dyadic neighbor relationships, rather than a more
detailed raster or vector GIS implementation, is justified on the basis that the
academic literature on mechanization suggests that two states neighboring each
other is the important factor, not more detailed geographic information (e.g.,
topographical features along the border). As such, this paper's modeling choice
captures the phenomenon of interest while avoiding the overhead of a full-blown
GIS implementation coupled with ABM
(e.g., \cite{harper2002modeling}).

@@

An important caveat is that the set of agents represented in the model does not
change over time: i.e., agents are the same from 1979 onwards. In real life,
of course, the agent population was not static over this time period, with many
states ceasing to exist in 1991 and new ones forming. Modeling these changes is
beyond the scope of this model. How a collapsed state's military assets are divvied up
among successor states is complex and not tractable to simple representation
(e.g. averaging). For example, the Soviet Union deployed most of its modern
military equipment in the far west of the country in anticipation of a conflict
with NATO, and so when the USSR collapsed Ukraine inherited the lion's share, leaving
little for successor states like Kazakhstan or Azerbaijan (\cite{roy1993military}). Such inheritance has
a strong effect on the mechanization of of successor states. Designing
complex and state-specific ``rules of inheritance'' was too much for this
initial effort but is an area for further development.

@@

\subsection{System-Level}

\subsection{Agent-Level}

@@

Learning ala \cite{bennett2006modelling}

>>>>

\begin{center}
	Word count: 4209
\end{center}
\section{Introduction}

@@

Agents exist in a spatial environment. Spatiality is represented by neighbor
relationships. Two states are defined as
being neighbors if they share a contiguous land border or a stretch of water
less than 400 miles. Data on neighbor relationships is taken from the Correlates
of War (COW) Direct Continuity dataset (\cite{cow_contiguity}).
Modeling spatiality as dyadic neighbor relationships, rather than a more
detailed raster or vector GIS implementation, is justified on the basis that the
academic literature on mechanization suggests that two states neighboring each
other is the important factor, not more detailed geographic information (e.g.,
topographical features along the border). As such, this paper's modeling choice
captures the phenomenon of interest while avoiding the overhead of a full-blown
GIS implementation coupled with ABM
(e.g., \cite{harper2002modeling} or \cite{bennett2006modelling}).

@@

An important caveat is that the set of agents represented in the model does not
change over time: i.e., agents are the same from 1979 onwards. In real life,
of course, the agent population was not static over this time period, with many
states ceasing to exist in 1991 and new ones forming. Modeling these changes is
beyond the scope of this model. How a collapsed state's military assets are divvied up
among successor states is complex and not tractable to simple representation
(e.g. averaging). For example, the Soviet Union deployed most of its modern
military equipment in the far west of the country in anticipation of a conflict
with NATO, and so when the USSR collapsed Ukraine inherited the lion's share, leaving
little for successor states such as Azerbaijan that were not important military
districts (\cite{roy1993military}). Such inheritance has
a strong effect on the mechanization of of successor states. Designing
complex and state-specific ``rules of inheritance'' was too much for this
initial effort but is an area for further development.

@@

At both the system- and agent-level, this agent-based model results in much
higher mechanization levels than occurred in the real world. First, let us
compare the average mechanization levels across all countries at each time-step
of the model. Although the set of agents is not exactly the same between the
model and reality (c.f. earlier issues with inheritance), this
gross comparison still is a useful measure of comparison. Model results are
the average of 100 runs, with all parameters (e.g.
\textit{EnemyMechSensitivity}) drawn from the distributions claimed by
\cite{sechser2010army}; empirical results are drawn from the National
Mechanization Index compiled by \cite{sechser2010army}.

\begin{table}[h]
	\centering
	\caption{Comparison}
	\begin{tabular}{|l r r|}
		\hline
		\textbf{Year} & \textbf{Model} & \textbf{Reality} \\
		1979 & 0.0156 & 0.0139 \\
		1981 & 0.0189 & 0.0167 \\
		1983 & 0.0229 & 0.0167 \\
		1985 & 0.0280 & 0.0183 \\
		1987 & 0.0351 & 0.0187 \\
		1989 & 0.0415 & 0.0197 \\
		1991 & 0.0512 & 0.0215 \\
		1993 & 0.0636 & 0.0229 \\
		1995 & 0.0804 & 0.0231 \\
		1997 & 0.1014 & 0.0270 \\
		1999 & 0.1319 & 0.0272 \\
		2001 & 0.1730 & 0.0291 \\
		\hline
	\end{tabular}
\end{table}

\newpage

Examining results at the agent-level reveals that the higher mechanization
levels of the model are much more pronounced for some agents than others. In
some cases, the model actually results in \textit{lower} mechanization levels,
with Mexico serving as one example.

\begin{table}[h]
	\centering
	\caption{Mexico}
	\begin{tabular}{|l r r|}
		\hline
		\textbf{Year} & \textbf{Model} & \textbf{Reality} \\
		1979 & 0.0013 & 0.0012 \\
		1981 & 0.0014 & 0.0012 \\
		1983 & 0.0018 & 0.0026 \\
		1985 & 0.0019 & 0.0013 \\
		1987 & 0.0021 & 0.0019 \\
		1989 & 0.0022 & 0.0027 \\
		1991 & 0.0024 & 0.0030 \\
		1993 & 0.0026 & 0.0033 \\
		1995 & 0.0030 & 0.0035 \\
		1997 & 0.0032 & 0.0074 \\
		1999 & 0.0036 & 0.0082 \\
		2001 & 0.0040 & 0.0075 \\
		\hline
	\end{tabular}
\end{table}

For some agents, however, mechanization levels are extremely high. 
Mechanization score is a percentage, and so values over 1 do not make sense, 
yet some countries end the simulation with mechanization scores over 1. Note
that one could easily right-bound values at 1, but this was not done in order to
see what values would be organically converged to.

\begin{table}[h]
	\centering
	\caption{Highest mechanization scores, 2001}
	\begin{tabular}{|l r|}
		\hline
		\textbf{Agent} & \textbf{Mechanization Score} \\
		Israel & 1.59 \\
		USA & 1.34 \\
		Kuwait & 1.14 \\
		Russia & 1.06 \\
		\hline
	\end{tabular}
\end{table}

There appears to be a correlation between number of enemies and high
mechanization score. For example, both the USA and Russia consistently had some
of the highest amount of enemies, peaking with 16 enemies for Russia in 1999
(this high number is attributable to the Kosovo crisis, when a confrontation between Russian and NATO forces
at Pristina Airport nearly led to war), and as per the above table are among the
most mechanized states. That Israel had many enemies during this time period is
obvious.

There also appears to be neighborhood effects. Mechanization is not evenly
distributed across the globe. Rather, it is geographically concentrated in a few
hot-spots. Agents belonging to the Middle East and North Africa (MENA) region
account for about 10\% of the overall agent population, yet make up 50\% of the
top 20 most mechanized agents at simulation end. Other regions
are correspondingly under-represented; for example, Asia does not have a single
state in the top 20.

\begin{table}[h]
	\centering
	\caption{Top 20 mechanized agents at simulation end, by region}
	\begin{tabular}{|l l l l l l|}
		\hline
		\textbf{MENA} & \textbf{Europe \& FSU} & \textbf{Africa} & \textbf{Americas} & \textbf{Asia} & \textbf{Oceania} \\
		Israel       & Russia         & Sudan        & USA    & (none) & (none) \\
		Kuwait       & France         & S. Africa    & Canada &        &        \\
		Iran         & UK             &              &        &        &        \\
		Libya        & Greece         &              &        &        &        \\
		Iraq         & Netherlands    &              &        &        &        \\
		Syria        & Poland         &              &        &        &        \\
		KSA          &                &              &        &        &        \\
		Egypt        &                &              &        &        &        \\
		Lebanon      &                &              &        &        &        \\
		Turkey       &                &              &        &        &        \\
		\hline
	\end{tabular}
\end{table}

@@

\subsection{Findings}

Two conclusions should be drawn. One, the model results in much higher
system-level mechanization than occurred in reality. Two, this higher
mechanization is not evenly distributed across agents; rather, a small number of
agents with many enemies and that all border each other become outliers and
hence drive the global average far upward. The explanation for both phenomena
lies in a positive feedback loop. Enemies and neighbors adjust their
mechanization scores in response to each other; hence, a geographic cluster of
enemies (e.g., the Middle East or Cold War-era Europe) will rapidly increase their
mechanization.

Three, because this model is a
faithful representation of the factors that \cite{sechser2010army} claim are
most important for mechanization, that model output fails to match reality
indicates that their claimed model is incomplete. There must be some ``brake''
on mechanization; put differently, some countervailing factor(s) must mitigate the
effect of the positive feedback loop identified above. This model cannot
conclusively identify what that ``brake'' is. The most plausible candidates are
factors examined by \cite{sechser2010army} and deemed statistically
insignificant such as regime type and economy, and factors not examined by
\cite{sechser2010army} at all such as culture.

This paper is a novel contribution to the literature. Although agent-based
models have found traction in the study of grievance, insurgency, and
other population-centric security phenomena (e.g., \cite{epstein2002modeling}),
they quite rarely are used to model international relations and state-centric security
phenomena despite some early work on this front (e.g., \cite{cederman1997emergent} and \cite{axelrod1997complexity}). 
This paper's agentization of an orthodox security
studies model thus is novel in method. Its findings also are important. They
show that they show a solely ``external'', security
environment-focused view of mechanization does not stand up to agentization.
Spatiality, inter-agent relations, autonomous agents with cognition, and positive feedback
loops---all hallmarks of a complex system---drive results that significantly
diverge from theoretical expectations of the orthodox view.

\subsection{Areas for Improvement}

Having taken a critical view of \cite{sechser2010army}, it only is fair that
this paper applies that lens to itself. There are two main areas for
improvement.

One, this model's representation of \cite{sechser2010army} could be tightened
up, as there currently is divergence between the two. For example, this model
examines a smaller agent set than their paper. Building ``inheritance''
processes to allow new states to emerge is desirable. This paper's
use of statistical coefficients from \cite{sechser2010army} also has some
methodological problems, a full description of which is omitted here due to
space constraints. It suffices to say that a better approach would be to conduct
more extensive parameter sweeps of relevant cognitive factors.

Two, this model could incorporate potential explanations of what the missing
``brake'' on mechanization might be. Economy and culture could be included in
the model. Doing so would allow exploration of whether these factors are
logically consistent and can act as a countervailing factor that controls the
positive feedback loop of mechanization. It cannot \textit{prove} that these or
any other factors actually exist in reality, but they can demonstrate how they
are or are not consistent in a simulated reality.

My hope is that making these changes will make this a publication-quality paper
capable of serving as one of two required papers for the comprehensive exams.

\newpage
